An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)

### 2022-04-07 manuscript/main.tex
rewrites

\title{An item response theory analysis of the matrix reasoning item bank (MaRs-IB)}
\shorttitle{IRT analysis of MaRs-IB}
\author{Samuel Zorowitz$^1$, Nathaniel D. Daw$^{1,2}$}
\affiliation{$^1$Princeton Neuroscience Institute, Princeton University, USA\\$^2$Department of Psychology, Princeton University, USA}

@@

Matrix reasoning tasks are one of the most well-studied and useful tasks in the behavioral sciences. In a typical administration, participants are presented with a 3x3 matrix in which 8 of 9 cells contains abstract figures or shapes. Participants are instructed to identify the missing figure from among several alternatives which are presented below the matrix. To do so, participants must use abstract reasoning to identify the piece that would complete the pattern. Matrix reasoning indexes fluid intelligence and is correlated with working memory capacity \citep{conway2002latent, kane2004generality, unsworth2005working, salthouse2014relations}. The progressive matrices has excellent predictive validity and is associated of academic performance \citep{roth2015intelligence} and college admission exams \citep{frey2004scholastic, koenig2008act}. In low-stakes testing environments, performance on the progressive matrices is associated with participant motivation and willingness to expend effort \citep{gignac2018moderate, gignac2019maximum}. For all of these reasons, the progressive matrices are increasingly used as a control measure in individual difference behavioral studies \citep{gillan2016characterizing, rouault2018psychiatric, moutoussis2021decision}.

Despite their utility, there are a number of challenges in using matrix reasoning measures in behavioral research. The first is the issue of copyright. The most prominent versions, including the Raven's progressive matrices \citep{raven2003raven} and the matrix subtest of the Wechsler Abbreviated Scale of Intelligence \citep{wechsler1999wechsler}, are not free to use and copyright may prevent these pen-and-paper tasks from being adapted into computerized tasks. A second challenge is those measures that are in the public domain typically have few items available. For example, the Hagen Matrices Test \citep{heydasch2014hagen} and International Cognitive Ability Resource \citep{condon2014international} have only 20 and 16 items available, respectively. The availability of only a limited number of items raises the issue of practice effects \citep{ng1974applicability, bors2003effect}, especially in the context of online experiments where multiple groups of researchers may be unknowingly administering the same items to the same participants.

In response to these challenges, \cite{chierchia2019matrix} developed the matrix reasoning item bank (MaRs-IB). The MaRs-IB is an open-source, free-to-use bank of 80 matrix reasoning items, where each item has three unique shape variants and two sets of distractors. The items were designed to vary considerably in their complexity (see Figure 1 for examples). In an initial validation study of a MaRs-IB short-form test, \cite{chierchia2019matrix} that measures of performance in a sample of 659 adults, adolescents, and children had excellent split-half and test-retest reliability. Moreover, performance on the MaRs-IB items were found to correlate with performance on the digit span tasks and ICAR matrix reasoning, indicating good convergent validity. Most importantly, the authors made available the percent correct data for each item (and variant) so that others researchers would be able to construct new MaRs-IB forms of custom difficulty and duration.

There are unfortunately two key obstacles preventing researchers from being fully able to do so, both of which stem from the design of the validation study. The original MaRs-IB short-form is a fixed-time and  fixed-order measure, where participants complete as many items as possible within 8-minutes. Because participants only completed 33 items on average, the first issue is that most items were completed by too few participants to establish reliable norms (an issue that is compounded when considering each item has six variants). Just over half the items (42/80) were completed by 100 or more participants. A second issue is that, because participants were free to choose for themselves whether to prioritize accuracy (number of correct responses) or productivity (number of items reached), it is likely that the percent correct data for items reached by fewer numbers of participants are biased. That is, items completed by fewer numbers of participants are more likely to have been completed by participants sacrificing accuracy for speed. Indeed, looking only at the easiest MaRs-IB items (dimension = 1), there is a strong positive correlation between number of completing participants and mean accuracy ($r = 0.673$).

Here we wish to overcome the limitations of the previous study to further validate and make useful the MaRs-IB. First, we collected a novel dataset of response to MaRs-IB items in a large sample of adults (N=1501). Using a fixed-length design with counterbalancing, we ensured that all items, and their variants, were completed by a suitable number of participants. Next we analyzed participants' response data hierarchical item response models, which allowed us to measure all items' difficulty and discrimination while controlling for participants' varying abilities. The item response models also allowed us to measure directly the extent to which that item variants are truly equivalent. We then demonstrate how these item response parameters can be used to construct new short-forms with desirable properties. Using these methods, we introduce three new MaRs-IB short-form tests with good internal consistency and convergent validity. 
All data and code, including items' estimated parameters and example code for test assembly, is available at: \url{https://github.com/szorowi1/mars-irt}.

@@

\subsection{Data collection}

A total of N=1584 participants were recruited from the Prolific Academic platform (https://www.prolific.co) to participate in an online behavioral experiment in July - August, 2021. Participants were eligible if they were 18 years or older and currently resided in the United States. This study was approved by the Institutional Review Board of Princeton University (\#7392), and all participants provided informed consent. Total study duration was approximately 6.4 minutes (2.4 sd) per participant. Participants received monetary compensation for their time (rate USD \$10/hr), plus an incentive-compatible bonus up to \$0.50 based on task performance (average payment: mean = \$1.30 [\$0.10 sd]). Monetary incentives have previously been shown to motivate good performance in low-stakes testing environments \citep{duckworth2011role, gignac2018moderate}.

After providing consent, participants completed a series of items from the MaRs-IB. The design of the MaRs-IB items have been described previously \citep{chierchia2019matrix}. Briefly, each MaRs-IB item consists of a 3 x 3 matrix. Eight of the nine resulting cells contain an abstract shape, while one cell on the bottom right-hand side of the matrix is empty. The participants' task was to complete the matrix by finding the missing shape among four possible alternatives. Items began with a 1200 ms fixation cross. Participants were then given up to 30 s to complete an item. After 25 s a clock appeared, indicating that 5 s remained before the next item began. An item ended when participants responded, or after 30 s had elapsed without response. Participants reviewed instructions before the task and were made to correctly complete three practice items before they could begin the task. Crucially, participants were instructed to respond carefully but to guess before the timer ran out. Our version of the MaRs-IB was based on publicly available code (\url{https://app.gorilla.sc/openmaterials/36164}).

In order to ensure adequate sampling of all items, and in contrast to the original MaRs-IB test form, all participants completed a pseudorandomly-selected set of 16 out of 64 possible MaRs-IB items \footnote{We did not collect new data for the 16 easiest items from the MaRs-IB (items 1, 2, 3, 4, 5, 7, 8, 9, 32, 33, 38, 41, 43, 48, 57, 68) as the original study or pilot testing found performance on these items to be at or near ceiling.}. Prior to data collection, we sorted the 64 items into 16 groups of four based on the number of changing relationships between the shapes of the matrix. Participants were then randomly assigned one item from each of the 16 groups. In this way, all participants received test forms of approximately equal difficulty. Importantly we also took steps to counterbalance the sampling of item variants, so that we had approximately the same number of responses available by item set (1, 2, or 3) and distractor type (minimal difference or or paired difference; see \cite{chierchia2019matrix} for details).

To ensure data quality, the data from multiple participants who completed the experiment were excluded prior to analysis for one or more of the following reasons: for rapid guessing -- defined here as responding in less than 3 s -- on 4 or more trials (N=70); for failing to respond on four or more trials (N=8); for minimizing or otherwise shrinking the browser window beyond what was required to view the stimuli (N=7). In total, 83 of 1584 (5.24\%) participants were excluded, leaving the data from N=1501 participants for analysis. Among these participants, each item (64 in total) was sampled approximately 375 times (sd = 12) and each item variant (384 in total) was sampled approximately 62 times (sd = 7).

\subsection{Descriptive statistics}

The majority of participants in the sample identified as women (men: N=670; women: N=811; non-binary or other: N=13; rather not say: N=2) and were 28.7 years old (sd = 9.9) on average. The sample was relatively well-educated, with the majority having completed a Bachelor's degree (N=507) or Master's degree or higher (N=322). By comparison, fewer participants endorsed having completed only some college (N=471), only a high school degree (N=199), or preferring not to say (N=2).

@@

Though not the primary focus of the current study, we briefly investigated the relationship between response accuracy and times. We used a mixed-effects model to predict participants' trial-level log-transformed response times as a function their response on that trial (correct or incorrect), their rest score (their total score on all other items), that item's difficulty (one minus the accuracy rate), and all possible interactions. We also specified an intercept per participant to control for each participants' overall speed.

The complete summary of results from the regression analysis are presented in Table XX. On average, participants spent 15.9 s per item (sd = 7.2 s, IQR = 10.1 - 21.5 s). The mixed-effects model identified significant main effects of accuracy ($\beta = 0.067, t = 9.84, p < 1e^{-6}$), rest score ($\beta = 0.125, t = 16.948, p < 1e^{-6}$), and item difficulty ($\beta = 0.137, t = 44.326, p < 1e^{-6}$). On average then, response times were slower for correct responses, more accurate participants, and more difficult items. There was also a significant positive accuracy by difficulty interaction (correct responses were even slower for more difficult items) and a significant negative accuracy by score interaction (correct responses were faster for better performing participants). Finally, there was a significant positive difficulty by score interaction, such that better performing participants exhibited slower responses for more difficult items. The three-way interaction term was not statistically significant.

In short, we corroborate the previous finding that item difficulty is associated with slower response times \citep{chierchia2019matrix}. In addition, we identified two notable patterns: better performing participants exhibited slower response times on average, which slowered further for more difficult times. This pattern of response time results is summarized in Figure \ref{fig:fig01}c. This pattern of results is consistent with controlled processing accounts, a point we return to in the Discussion.

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{figures/fig01.png}
\caption{\label{fig:fig01}This is a figure caption.}
\end{figure}

Summary paragraph?

\subsection{Item response models}

To measure the psychometric properties of each item, we used item response models. Given the hierarchical nature of the MaRs-IB -- 64 items with six variants each -- we specifically modeled participants' response data using item using a series of item cloning models \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. In item cloning models, the variants of an item (or "clones") are explicitly model as having been generated from the same template item (or "family"). The basis of all of the models used here is the three-parameter logistic (3PL) item response model, where the probability of a correct response ($y_{ijk} = 1$) for person $i$ on item clone $k$ belonging to item family $j$ is:

\begin{equation}
p(y_{ijk} = 1) = \gamma_{jk} + (1-\gamma_{jk}) \cdot \text{logit}^{-1} \left( \alpha_{jk} \cdot \theta_i - \beta_{jk} \right)
\end{equation}

\noindent where $\theta_i$ is the latent ability for person $i$, and $\beta_{jk}$, $\alpha_{jk}$, and $\gamma_{jk}$ are the difficulty, discrimination, and guessing parameters for item clone $k$ of item family $j$. Since guessing parameters are challenging to estimate in the absence of very large amounts of data \citep{han2012fixing}, we fixed the guessing parameter for each item clone to the nominal guessing rate ($\gamma_{jk} = 0.25$).

To account for the nested structure of the MaRs-IB, the item difficulty and discrimination parameters were estimated using an additive multilevel item structure model \citep{cho2014additive}. Specifically, the difficulty of item clone $k$ belonging to item family $j$ was expressed as:

\noindent Every difficulty parameter is thus modeled as an intercept ($\mu_\beta$), reflecting the average difficulty across all items, and four components:

\item $\sum_{n=1}^N Q_{jn} \delta_{\beta n}$: the effect of the level-1 attributes, where $Q_{jn}$ is the value of attribute $n$ for item family $j$, and $\delta_{\beta n}$ is the effect of attribute $n$. This component is the fixed effects contribution of item family to item difficulty.

\item $\epsilon_{\beta j}$: the level-1 residual, or variance unexplained by the item family attributes. This component is the random effects contribution of item family to item difficulty.

\item $\sum_{m=1}^M R_{km} \delta_{\beta m}$: the effect of the level-2 attributes, where $R_{km}$ is the value of attribute $m$ for item clone $k$, and $\delta_{\beta m}$ is the effect of attribute $m$. This component is the fixed effects contribution of item clone to item difficulty.

\item $\epsilon_{\beta k}$: the level-2 residual, or variance unexplained by the item clone attributes. This component is the random effects contribution of item clone to item difficulty.

To elaborate, level-1 attributes describe the properties of an item family. Here we model the effects of two properties: feature number, or the number of unique shapes in each cell of an item; and rule number, or the number of changing relationships among the shapes (e.g. color change, shape change, position change) a participant must consider in order to solve the item.

In turn, level-2 attributes describe properties of an item clone. Here we model the effects of two additional properties: distractor type, or whether the distractors were generated following the minimal difference (MD) strategy and a paired difference (PD) strategy; and mean response time, or the average time spent on that item. Mean response time was orthogonalized with respect to the three other item attributes, and therefore reflects any residual structure after accounting for all other modeled effects.

Similarly, item discrimination parameters were also expressed as the sum of an equivalent set of components:

\noindent The fixed effect components of the item discrimination parameters included the same level-1 and level-2 attributes as for the item difficulty parameters (i.e. number of features, number of rules, distractor type, average response time). During estimation, the item discrimination parameters were restricted to be in the range $\alpha_{jk} \in [0, 5]$.

>>>>

% override apa6 section headers
% https://tex.stackexchange.com/questions/125537/how-to-modify-subsubsection-header-apa6-cls
\makeatletter
\renewcommand{\subsubsection}{\@startsection{subsubsection}{3}
  {\z@}%
  {\b@level@two@skip}{\e@level@two@skip}%
  {\normalfont\normalsize\bfseries}}
\makeatother

% override apa6 first paragraph index
% https://tex.stackexchange.com/questions/155028/indentation-apa6-class-for-2nd-paragraphs-after-title
% \makeatletter
%   \b@level@one@skip=-2.5ex plus -1ex minus -.2ex
%   \b@level@two@skip=-2.5ex plus -1ex minus -.2ex
% \makeatother

\title{An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)}
\shorttitle{IRT analysis of MaRs-IB}
\author{Samuel Zorowitz$^1$, Nathaniel D. Daw$^{1,2}$}
\affiliation{$^1$Princeton Neuroscience Institute, Princeton University, USA\\$^2$Department of Psychology, Princeton University, USA}

@@

Matrix reasoning tasks are among the most well-studied and useful measures in the behavioral sciences. Matrix reasoning requires participants to use abstract reasoning to solve puzzles. Matrix reasoning indexes fluid intelligence and is correlated with working memory capacity \citep{conway2002latent, kane2004generality, unsworth2005working, salthouse2014relations}. In low-stakes testing environments, performance on the progressive matrices is associated with participant motivation and willingness to expend effort \citep{gignac2018moderate, gignac2019maximum}. The progressive matrices has excellent external validity, and is predictive of academic performance \citep{roth2015intelligence} and scores on college admission exams \citep{frey2004scholastic, koenig2008act}. % Thus, matrix reasoning tasks are increasingly utilized as control measure in individual difference behavioral studies \citep{gillan2016characterizing, rouault2018psychiatric, moutoussis2021decision}.

Unfortunately there are a number of challenges in using matrix reasoning tasks as part of behavioral research. One challenge is the problem of copyright. The most prominent matrix reasoning tasks, such as the Raven's progressive matrices \citep{raven2003raven} and the WAIS matrix subtest \citep{wechsler1999wechsler}, are typically not free to use and have restrictions against being computerized. A second challenge is that, across the measures in the public domain, there are relatively few items available. For example, the Hagen Matrices Test \citep{heydasch2014hagen} and ICAR matrix reasoning test \citep{condon2014international} have only 20 and 16 items, respectively. The availability of only a limited number of items raises the possibility of practice effects \citep{ng1974applicability, bors2003effect}, especially in the age of online experiments where multiple groups of researchers may be unwittingly administering the same test to many of the same participants.

In response to these challenges, \cite{chierchia2019matrix} developed, and released into the public domain, the matrix reasoning item bank (MaRs-IB). The MaRs-IB includes 80 matrix reasoning puzzles. Each puzzle consists of a 3x3 matrix containing abstract shapes in eight out of nine cells. To solve a puzzle, participants must deduce which of four response options completes the matrix based on the pattern of changing relations among the shapes across cells (Figure 1). The puzzles in the MaRs-IB vary considerably in their complexity, both in the number of shapes per cell and in the number of changing relations across shapes. Importantly each template puzzle has six clones, which are matched in terms of complexity but differ in their perceptual properties (one of three shape sets) and their response options (one of two distractor sets). Thus, the MaRs-IB offer considerable utility insofar that they are free to use, have many items, and many unique items to prevent practice effects.

A preliminary investigation of the psychometric properties of the MaRs-IB items proved promising. Indeed, \cite{chierchia2019matrix} found that scores on a short-form matrix reasoning test using MaRs-IB items had good split-half reliability and test-retest reliability. The authors also found that performance observed scores on this test were moderately correlated with a digit span task and the ICAR matrix reasoning test, indicating good convergent validity. % note the release of summary statistics here?

Unfortunately, there are two notable limitations in this initial investigation. The first is that, due to the design of the matrix reasoning test, there is little data available for the majority of the items in the MaRs-IB. Just over half the items (42/80) were completed by 100 or more participants. As such, there is too little data to characterize the psychometric properties of each item with reasonable confidence (an issue that is exacerbated when considering each template item has six variants). A second issue is that, also due to the design of the matrix reasoning test, participants were allowed to choose for themselves whether to prioritize accuracy (number of correct responses) or productivity (number of items reached). As a consequence, items later in the sequence are likely to be contaminated by participants sacrificing accuracy for speed, thereby distorting the summary statistics for these items.\footnote{appendix: Indeed, looking only at the easiest MaRs-IB items (dimension = 1), there is a strong positive correlation between number of completing participants and mean accuracy ($r = 0.673$).} Thus, caution is warranted in interpreting the results of the original study and additional validation is warranted.

%Better intro sentence tk: mention that this is a more extensive psychometric validation of the MaRs-IB in a much larger dataset, with three major improvements.
The most important contribution of the current work is the evaluation of the psychometric properties of (almost) every item in the MaRs-IB using item response theory (IRT) \citep{embretson2013item, de2013theory}. IRT models confer several important benefits for the purposes of psychometric validation. First they provide an estimate of the difficulty and discrimination of each item, where the latter is an index of how informative or reliable an item is. Through these parameters, IRT models allows for the computation of the test information function (TIF), which measures how sensitive a test is at measuring performance at varying levels of ability. Finally, IRT make possible optimal test assembly \citep{van1998optimal}, or design of new tests with maximal sensitivity to measuring performance in a specified range of abilities under varying constraints (e.g. test length, time limits, parallel forms with equal TIF).

A second contribution offered here is an analysis of the relationships between attributes of MaRs-IB items and their psychometric properties using explanatory item response models \citep{de2004explanatory, wilson2008explanatory}. Explanatory IRT models provide a framework for measuring how item parameters (e.g. difficulty) are shaped by their features. As such, these models provide means to evaluate how successful was the generation of the items in the MaRs-IB. For example, we can quantify the extent to which features of item complexity (e.g. number of elements, number of changing relations) predict item difficulty. Moreover, we can determine whether two items that vary only in their distractor set are psychometrically equivalent. Explanatory IRT models have the additional potential benefit of producing more accurate estimates of item parameters \citep{neuhaus2006separating}.

%item clones would be isomorphic (psychometrically equivalent). In any realistic situation, the 2P-F is misspecified as it ignores the psychometric deviations introduced in the item cloning process (Glas & van der Linden, 2003).

A third contribution is the decomposition of variance in item functioning from modeled and unmodeled sources using additive multilevel item structure (AMIS) models \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. This framework provides another means by which to evaluate the success of the design of the MaRs-IB by quantifying the residual variance in item parameters after accounting for item attributes. For example, two items of equal complexity may turn out not be psychometrically equivalent if unmodeled attributes (e.g. the types of changing relations) implicitly impact item functioning. Similarly, two item clones may not be psychometrically equivalent due to differences in their superficial properties. Quantifying the degree of error in the item cloning process is critical, as the degree of nonequivalent dictates whether clones are exchangeable. That is, if clones are not psychometrically equivalent, then they cannot be used interchangeably in parallel short forms.

The paper proceeds as follows. First we analyze participants' response data hierarchical item response models, which allowed us to measure all items' difficulty and discrimination while controlling for participants' varying abilities. The item response models also allowed us to measure directly the extent to which that item variants are truly equivalent. We then demonstrate how these item response parameters can be used to construct new short-forms with desirable properties. Using these methods, we introduce three new MaRs-IB short-form tests with good internal consistency and convergent validity.

All data and code, including items' estimated parameters and example code for test assembly, is available at: \url{https://github.com/szorowi1/mars-irt}.

@@

\subsection{Objectives}

The purpose of the first study was to calibrate the items of the MaRs-IB, using response data collected from a large number of adult participants from the general population. In particular, we sought to accomplish the following three objectives: (1) to characterize the psychometric properties of the items in the MaRs-IB using item response models; (2) to investigate the associations between item attributes and their psychometric properties, and (3) to determine the equivalence of items with similar complexities, and clones of the same item templates.

\subsection{Methods}

\subsubsection{Participants}

A total of N=1584 participants were recruited from the Prolific Academic platform (\url{https://www.prolific.co}) to participate in an online experiment between July - August, 2021. Participants were eligible if they were 18 years or older and lived in the United States. Total study duration was approximately 6.4 minutes (sd = 2.4) per participant. Participants received monetary compensation for their time (rate: \$10 USD/hr), plus a performance-based bonus up to \$0.50 USD. On average, participants earned a total of \$1.30 USD (sd = \$0.10). We offered performance bonuses as they have been found to motivate performance in low-stakes testing environments \citep{duckworth2011role, gignac2018moderate}. This study was approved by the Institutional Review Board of Princeton University (\#7392), and all participants provided informed consent.

To ensure data quality, the data from multiple participants were excluded prior to analysis (see Exclusion Criteria below) leaving the data from N=1501 participants for analysis. The majority of participants in this sample identified as women (men: N=670; women: N=811; non-binary or other: N=13; rather not say: N=2) and were 28.7 years old (sd = 9.9) on average. The sample was relatively well-educated, with the majority having completed a Bachelor's degree (N=507) or Master's degree or higher (N=322). By comparison, fewer participants endorsed having completed only some college (N=471), only a high school degree (N=199), or preferred not to say (N=2).

\subsubsection{Procedure}

After providing consent, participants completed 16 items from the MaRs-IB. The design of the MaRs-IB items have been described previously \citep{chierchia2019matrix}. Briefly, each MaRs-IB item consists of a 3 x 3 matrix. Eight of the nine resulting cells contain abstract shapes, while one cell on the bottom right-hand side of the matrix is empty. Participants are instructed to ``complete the matrix'' by finding the missing shape from among four possible alternatives.

Each trial of the experiment began with a fixation cross, lasting  1200 ms, after which a MaRs-IB item was presented. Participants were then given up to 30 s to solve the puzzle. After 25 s a clock appeared, indicating that 5 s remained before presentation of the next item. A trial ended when participants responded, or after 30 s had elapsed without response. Before the trials began, participants reviewed instructions and were made to correctly complete three practice items. Participants were told to respond carefully but to guess before the timer ran out. The format of the experiment and instructions were adapted from code made publicly available by the original authors (\url{https://app.gorilla.sc/openmaterials/36164}).

In order to ensure sufficient and uniform sampling of the MaRs-IB items, all participants completed 16 items pseudorandomly selected from the 64 available MaRs-IB items.\footnote{We did not collect new data for the 16 easiest items from the MaRs-IB (items 1, 2, 3, 4, 5, 7, 8, 9, 32, 33, 38, 41, 43, 48, 57, 68) as the original study or pilot testing found performance on these items to be at or near ceiling.} The item sets were constructed as follows: First, we subdivided the 64 items into 16 sets of four based on their complexity. Participants were then randomly assigned one item from each of the 16 sets. In this way, all participants received test forms of approximately equal difficulty. We also counterbalanced which variant of each item was presented across participants, such that we had approximately the same number of responses available by shape set (1, 2, or 3) and distractor type (minimal difference or or paired difference; see \cite{chierchia2019matrix} for details).

\subsubsection{Exclusion criteria}

The data from multiple participants who completed the experiment were excluded prior to analysis for one or more of the following reasons: rapid guessing (responding in less than 3 s) on four or more trials (N=70); failing to respond on four or more trials (N=8); or minimizing the browser window to a size smaller than the minimally required dimensions (N=7). In total 83 of 1584 (5.2\%) participants were excluded, leaving the data from N=1501 participants available for analysis. Among these participants, each item template (64 in total) was sampled approximately 375 times (sd = 12) and each item clone (384 in total) was sampled approximately 62 times (sd = 7).

\subsubsection{Item response models}

We employed item response models to characterize the psychometric properties of each item. Specifically, we used additive multilevel item structure (AMIS) models with random residuals \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. In item structure models, item clones are modeled as descending from a template item. The basis of all of these models was the three-parameter logistic (3PL) item response model, where the probability of a correct response ($y_{ijk} = 1$) for person $i$ on item clone $k$ belonging to item family $j$ is:

\begin{equation} \label{eq:1}
p(y_{ijk} = 1) = \gamma_{jk} + (1-\gamma_{jk}) \cdot \text{logit}^{-1} \left( \alpha_{jk} \cdot \theta_i - \beta_{jk} \right)
\end{equation}

\noindent where $\theta_i$ is the latent ability for person $i$, and $\beta_{jk}$, $\alpha_{jk}$, and $\gamma_{jk}$ are the difficulty, discrimination, and guessing parameters for item clone $k$ of item family $j$. Because guessing parameters are challenging to reliably estimate in the absence of very large amounts of data \citep{han2012fixing}, we fixed the guessing parameter for each item clone to the nominal guessing rate ($\gamma_{jk} = 0.25$).

The difficulty and discrimination parameters were estimated following an additive multilevel item structure. Concretely, the difficulty of item clone $k$ belonging to item template $j$ was expressed as:

\noindent To elaborate, each item difficulty parameter was modeled as an intercept ($\mu_\beta$), reflecting the average difficulty across all items, and four additional components:

\item $\sum_{n=1}^N Q_{jn} \delta_{\beta n}$: the effect of the template (level-1) attributes on item difficulty, where $Q_{jn}$ is the value of attribute $n$ for item template $j$, and $\delta_{\beta n}$ is the effect of attribute $n$.

\item $\epsilon_{\beta j}$: the template (level-1) residual, or residual variability in item difficulty of the item template unexplained by its attributes.

\item $\sum_{m=1}^M R_{km} \delta_{\beta m}$: the effect of the clone (level-2) attributes on item difficulty, where $R_{km}$ is the value of attribute $m$ for item clone $k$, and $\delta_{\beta m}$ is the effect of attribute $m$.

\item $\epsilon_{\beta k}$: the clone (level-2) residual, or residual variability in item difficulty of the item clone unexplained by its attributes.

Similarly, item discrimination parameters were also modeled as the sum of an equivalent set of components:

\noindent where the interpretation of each component is the same as for item difficulty.

The AMIS model framework provides a straightforward means of achieving Aims 2 and 3. Via the attribute effect components, we can test for any associations between item attributes and parameters (Aim 2). So too, via the error terms, we can quantify the magnitude of residual variance in item parameters from unmodeled sources both at the level of item templates and item clones (Aim 3). If the residual variance in item parameters across item clones is sufficiently large, then we can conclude item clones are neither psychometrically equivalent nor exchangeable.

In the following analyses, we consider two template (level-1) attributes: feature number, or the number of unique shapes in each cell of an item; and rule number, or the number of changing relationships among the shapes (e.g. color change, shape change, position change) a participant must consider in order to solve the item. We predicted that both attributes would be positively related to item difficulty. In contrast, we had no hypotheses about the relationship between these attributes and item discrimination.

In turn, we consider two clone (level-2) attributes: distractor type, or MD-type distractors versus PD-type distractors; and mean response time, or the average time spent deliberating on that item. Based on the preliminary findings of \cite{chierchia2019matrix}, we predicted there would be no association between distractor type and either type of item parameter. We note that, in contrast to all of the other modeled attributes, average response time is not an intrinsic property of an item. We opted to include it in our models, however, because of its previously established relationship with item difficulty \citep{chierchia2019matrix}, and because including response time as a covariate has previously been shown to improve parameter estimation \citep{bertling2018using}. To prevent any collinearity between the attributes, mean response time was orthogonalized with respect to the three other item attributes, and therefore reflects any residual structure after accounting for all other modeled effects.\footnote{Note that we intentionally do not include shape set, or the superficial properties of an item, as covariates in our model. This is due to the would-be complexity of doing so. Though each item has three clones, the clones do not all have the same perceptual patterns. There are XX unique patterns. Future studies might consider even larger datasets to study perceptual influence.}

\begin{equation} \label{eq:4}
\theta_i = \sum_{p=1}^P X_{ip} \rho_p + \epsilon_i    
\end{equation}

\subsubsection{Model Fitting}

All models were estimated within a Bayesian framework using Hamiltonian Monte Carlo as implemented in Stan (v2.22) \citep{carpenter2017stan}. For all models, four separate chains with randomised start values each took 7500 samples from the posterior. The first 5,000 samples from each chain were discarded. As such, 10,000 post-warmup samples from the joint posterior were retained. The $\hat{R}$ values for all parameters were equal to or less than 1.01, indicating acceptable convergence between chains, and there were no divergent transitions in any chain.

During estimation, the item discrimination parameters were restricted to be in the range $\alpha_{jk} \in [0, 5]$.

\subsubsection{Model comparison}

The goodness of fit of the models was compared using Bayesian leave-one out cross-validation \citep{vehtari2017practical}, which has been found to perform better than more traditional information criteria for comparing item response models \citep{luo2017performances}. Importantly, we computed the conditional leave-one-cluster out (LOCO) cross- validation \citep{merkle2019bayesian}, which measures a model's ability to generalize to held out items (rather than held out responses or held out participants) --- i.e., for inferring generalization to additional items constructed using the same features rather than to additional participants sampled from the same population.

\subsubsection{Goodness of fit}

The fit of the best-fitting model to the data was evaluated through posterior predictive model checks \citep{gelman1996posterior, levy2017bayesian}. Item fit was assessed using the Orlando and Thissen index ($S-\chi^2$) \citep{toribio2011discrepancy}, which compares the observed and expected proportions of participants who answer an item correctly at each raw-score group. The unidimensionality of the data was assessed using the generalized dimensionality discrepancy measure (GDDM) \citep{levy2011generalized}, which tests for violation of the assumption of local independence. This latter measure is crucial for ensuring there are negligible levels of local item dependence in the data, which may otherwise lower the efficiency of future test forms due to the inclusion of correlated, redundant items.

\subsection{Results}

\subsubsection{Descriptive statistics}

@@

We investigated the relationship between response accuracy and times using a random-intercepts linear regression model. Participants' log-transformed, item-level response times were predicted as a function of item accuracy, rest score (observed score for all other items), and item difficulty (one minus the proportion correct on that item). The results are summarized in Table XX. On average, participants spent 15.9 s (sd = 7.2 s) on each item. There were significant main effects of accuracy ($\beta = 0.067$, $t = 9.84$, $p < 0.001$), rest score ($\beta = 0.125$, $t = 16.95$, $p < 0.001$), and item difficulty ($\beta = 0.137, t = 44.33, p < 0.001$) such that response times were slower for correct responses, better-performing participants, and more difficult items.

There was also a significant accuracy by difficulty interaction (correct responses were even slower for more difficult items), and a significant accuracy by score interaction (correct responses were faster for better performing participants). Finally, there was a significant difficulty by score interaction, such that better performing participants exhibited slower responses for more difficult items. This pattern of response time results is visible in Figure \ref{fig:fig01}c. The three-way interaction term was not statistically significant.

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{figures/fig01.png}
\caption{\label{fig:fig01}Summary of performance on the MaRs-IB items. (A) The distribution of total scores across all 1501 participants. (B) The distribution of proportion correct responses across all 384 items. (C) The distribution of participants' median response times across items broken down by participants' total scores and item difficulty (one minus proportion correct responses). Error bars indicate bootstrapped 95\% confidence intervals around the mean.}
\end{figure}

\subsubsection{Alternative item structure models}

We fit a series of nested IRT models in order to identify the model that best explained participants' response data. The models varied in how item parameters were estimated, with successive models allowing for greater variability in parameters across item templates and clones. Identifying the model that most accurately captures the structure of item parameters is important in the context of item cloning, as improperly modeled variance can result in the biased estimation of item parameters \citep{lathrop2017item}.

We fit the models in two waves. The first wave was comprised of four models in which the item discrimination parameters were fixed ($\alpha = 1$).\footnote{Note that these models still included a fixed guessing parameter ($\gamma = 0.25$). We also estimated models with no guessing parameters (i.e. one-parameter logistic or Rasch models), but model comparison indicated these models were worse fits to the data (Table SXX).} We did this in order to minimize the number of total models that we would otherwise need to fit (i.e. all permutations of the four models below, applied to item difficulty and discrimination parameters independently). The first four models were as follows:

\begin{itemize}

\item Model 1a: Item difficulty is a function only of template attributes (number of shapes, number of changing relations). In this model, items of equal complexity are psychometrically equivalent as are all item clones.

\item Model 1b: Item difficulty is a function of template attributes and residuals. In this model, items of equal complexity may not be equally difficult, but all clones of an item are still psychometrically equivalent.

\item Model 1c: Item difficulty is a function of both template and clone attributes (including distractor type and average response time), and template-level residuals. In this model, item clones may systematically vary in their difficulty as a function of their attributes.

\item Model 1d: Item difficulty is function of both template/clone attributes and residuals. In this model, neither item templates nor clones are psychometrically equivalent. The degree of nonequivalence is dependent on the magnitude of the residual terms.

\end{itemize}

The goodness-of-fit of the four models is summarized in Table \ref{table:1}. Model 1d demonstrated a better fit of the data than Model 1a ($\Delta$ LOCO = 1223.69, se = 44.94), Model 1b ($\Delta$ LOCO 511.58, se = 28.77), and Model 1c ($\Delta$ LOCO = 187.96, se = 17.25). Thus, Model 1d was chosen as the starting point for the second wave of model fitting.

Next we estimated a second set of three models in which the item discrimination parameter was a free parameter. For all models, the item difficulty parameters were specified according to Model 1d. The structure of the three models were as follows:

\begin{itemize}

    \item Model 2a: Item discrimination is a function only of both template and clone attributes. Here we include all attributes at once having found that they were all predictive of item difficulty (see below).

    \item Model 2b: Item discrimination is a function of both item attributes and template residuals. In this model, items of equal complexity may not be equally discriminating.

    \item Model 2c: the item discrimination parameters were modeled with fixed and random effects at both levels. In this model, neither item templates nor clones are equivalently discriminating. The degree of nonequivalence is dependent on the magnitude of the residual terms.

\end{itemize}

The goodness-of-fit of these models is summarized in Table \ref{table:1}. Notably, we did not find the most complex model (Model 2c) to be the best-fitting model. Although Model 2c was preferred to Model 2a ($\Delta$ LOCO = 2.13, se = 2.04), it was not preferred to Model 2b ($\Delta$ LOCO = -1.44, se = 0.85).
%We note that the authors of the LOO metric recommend that standard errors be at least twice as large as the difference between models to be confident that one model truly provides a better fit.
However, these differences in LOCO values from each model to the next were each within two standard errors of the mean, indicating only weak predictive improvement of Model 2b over the others \citep{vehtari2022cv}. As such we proceed by select Model 2b as the best-fitting model overall, but proceed in discussing it with caution. We also note that Model 2b fit clearly better than Model 1d ($\Delta$ LOCO = 85.91, se = 6.33, supporting the inclusion of varying item discrimination parameters).

\subsubsection{Objective 1}
